PTFO 8-8695: Two Stars, Two Signals, No Planet

### 2020-06-16 paper/ms.tex
update \explain environment calls

\explain{All paragraphs in this section have been heavily reworded to
emphasize RUWE over $\chi_{\rm red}^2$, as requested by the referee.}
A separate line of evidence for binarity is the Gaia DR2 astrometry.
As noted in Section~\ref{sec:observations}, the Gaia DR2 astrometric
solution for \ptfo\ shows a 10.3$\sigma$ ``astrometric excess'', a
parameter that quantifies the degree to which a single-star model
fails to fit the astrometric measurements.  Specifically, the
single-source astrometric model yielded $\chi^2=325.2$.  There are 121
astrometric measurements, and 5 free parameters, and therefore 116
degrees of freedom. The reduced $\chi^2$ is 2.80.  While the majority
of stars with comparable brightness in Gaia do not show such poor
goodness-of-fit\added{, many M dwarfs do, because of a color-dependent
term in the mission's PSF model} \citep[see][Appendix
A]{lindegren_gaiasoln_2018}.  The reduced $\chi^2$ has been
``renormalized'' to account for this, yielding a modified statistic
called the renormalized unit weight error (RUWE\footnote{ See the Gaia
DPAC technical note GAIA-C3-TN-LU-LL-124-01,
\url{http://www.rssd.esa.int/doc_fetch.php?id=3757412}, accessed
2020-04-27. }).

@@

\explain{Note: Citations to Parks et al 2014 and Stauffer et al 2015
were added to the paragraph below} When physical explanations are not
forthcoming, we often resort to taxonomy.  By searching the
literature, we found about a dozen light curves with similar
morphologies to \ptfo, drawn from surveys of low-mass weak-lined T
Tauri stars in regions including $\rho$ Oph, Upper~Sco, Taurus, and
perhaps the Pleiades
\citep{parks_periodic_2014,rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
The clearest matches came from K2 \citep{howell_k2_2014}, but a few
analogs were potentially also found using 2MASS (YLW 1C and YLW 10C;
\citealt{parks_periodic_2014}) and CoRoT
(\citealt{stauffer_csi_2015}).  We downloaded a subset of the K2 light
curves from MAST, opting for the EVEREST reductions
\citep{luger_everest_2016,luger_update_2018}.  They are plotted in
Figure~\ref{fig:brethren}.

\explain{Note: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars\added{ (under the
    definition of {\it e.g.}, \citealt{fang_star_2009})}.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. } 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.
\replaced{This is the first connection of \ptfo\ with this class of
objects, likely because the TESS data enabled us to put the dips in
the context of the asymmetric out-of-dip modulation.} {While this was
in fact mentioned by \citet{stauffer_orbiting_2017}, the
asymmetric out-of-dip modulation shown by the TESS data solidifies the
connection.}

@@

\explain{Note :Removed ``high-latitude accretion hotspots'' from the
list in this paragraph}
The key question is what causes the transient dips. This is an
unsolved problem not only for \ptfo\ but also for the emerging class
of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations
are that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
{\it (iv)} high-latitude starspots;
or {\it (v)} dust clouds of \replaced{plausible
composition}{particular compositions
\citep[][Figure~10]{zhan_complex_2019}}.  We also view the possibility
of {\it (vi)} tidally disrupted planetary or cometary material to be
implausible, given the synchronicity between dip and rotation periods
seen across many systems.  \added{Finally, {\it (vii)} transits of an
enshrouded protoplanet can be added to the list of disfavored
explanations, based on the chromaticity, asymmetric phase curve, and
light-curve changes seen in \ptfo.}

\explain{Note: ``High-latitude accretion hotspots'' was moved to this
paragraph.}
The explanations that we view as not yet ruled out include {\it A)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation
radius; {\it B)} occultations of \replaced{starspots}{hotspots on the
stellar surface} by an optically thick disk \citep{zhan_complex_2019};
and {\it C)} high-latitude accretion hotspots passing behind the star
\citep{yu_tests_2015}.  \replaced{The evolution of \ptfo\
over}{Observations of \ptfo\ from} the past decade may offer some
hints.  \deleted{Specifically, }\ptfo's transition between having
none, one, and multiple dips per cycle seems important
\added{(Figure~\ref{fig:o_minus_c})}.  \deleted{The dips would then
need to be caused by material that was somehow disrupted from the
planet, but somehow remained co-orbital for an extended duration. This
seems implausible.} \added{ The H$\alpha$ excess observed by
\citet{johnskrull_h_2016} also seems like an important piece of
evidence.  }

\explain{Note: This paragraph is newly added, to support our view of the
high-latitude accretion hotspot being plausible for \ptfo.}
The high-latitude accretion hotspot scenario seems qualitatively
capable of explaining most of the characteristics of \ptfo, including
the fact that its H$\alpha$ equivalent width of 11.4$\,$\AA\
\citep{van_eyken_ptf_2012} places it near the ``classical/weak'' T
Tauri division \citep{briceno_cidaII_2019}, and the fact that its
excess H$\alpha$ component varies in velocity between $-200$ and
$+200$ km$\,$s$^{-1}$ \citep{johnskrull_h_2016}.  The lack of a
NIR-excess could be problematic, but does not eliminate the
possibility that a tenuous amount of dust or gas remains near the
star, and is able to accrete onto it.  We view the main problem with
this explanation to be the likelihood that it explains the population
of transient flux dips.  RIK-210 in particular has dips too deep
($\sim$20\%) and too asymmetric to be explained by occultations of a
hot spot \citep{david_transient_2017}.

\explain{Note: This paragraph has been expanded to clarify the
possibility of a hotspot being occulted by a disk.}
Explanations {\it A} and {\it B} (circumstellar material at the
co-rotation radius, and occultations of spots by a disk) have the
appeal that they are flexible enough to explain not only the transient
and persistent-dip stars, but also the ``scallop shells''
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not fall
along the line of sight.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

@@

\explain{Updated this paragraph to remove spot occultations by a disk as
a ``preferred explanation''.}
The physical mechanism that explains the transient and persistent flux
dips is unresolved. Our preferred explanation is that a clumpy torus
of dust or gas is near the Keplerian co-rotation radius, and
periodically occults the star.  The \added{chromaticity,
out-of-eclipse variability, and} jumping orbital phase
\replaced{disfavored}{of \ptfo\ all disfavor} the explanation of an
enshrouded, transiting protoplanet.  Though \ptfo\ may not be a
planet,\deleted{ as we and others had hoped,} understanding it and its
analogs is a worthy problem\replaced{.}{ that seems likely to provide
insight into how gas and dust disperse from young protostellar disks.}
\replaced{It might even}{Understanding this process would, in turn,}
teach us about the birth environments of the majority of
habitable-zone Earth-sized planets \citep{dressing_occurrence_2013}.

>>>>

\explain{NOTE: All paragraphs in this section have been heavily reworded to
emphasize RUWE over $\chi_{\rm red}^2$, as requested by the referee.}
A separate line of evidence for binarity is the Gaia DR2 astrometry.
As noted in Section~\ref{sec:observations}, the Gaia DR2 astrometric
solution for \ptfo\ shows a 10.3$\sigma$ ``astrometric excess'', a
parameter that quantifies the degree to which a single-star model
fails to fit the astrometric measurements.  Specifically, the
single-source astrometric model yielded $\chi^2=325.2$.  There are 121
astrometric measurements, and 5 free parameters, and therefore 116
degrees of freedom. The reduced $\chi^2$ is 2.80.  While the majority
of stars with comparable brightness in Gaia do not show such poor
goodness-of-fit\added{, many M dwarfs do, because of a color-dependent
term in the mission's PSF model} \citep[see][Appendix
A]{lindegren_gaiasoln_2018}.  The reduced $\chi^2$ has been
``renormalized'' to account for this, yielding a modified statistic
called the renormalized unit weight error (RUWE\footnote{ See the Gaia
DPAC technical note GAIA-C3-TN-LU-LL-124-01,
\url{http://www.rssd.esa.int/doc_fetch.php?id=3757412}, accessed
2020-04-27. }).

@@

\explain{NOTE: Citations to Parks et al 2014 and Stauffer et al 2015
were added to the paragraph below} When physical explanations are not
forthcoming, we often resort to taxonomy.  By searching the
literature, we found about a dozen light curves with similar
morphologies to \ptfo, drawn from surveys of low-mass weak-lined T
Tauri stars in regions including $\rho$ Oph, Upper~Sco, Taurus, and
perhaps the Pleiades
\citep{parks_periodic_2014,rebull_rotation_2016,david_transient_2017,stauffer_orbiting_2017,stauffer_rotevol_2018,rebull_usco_2018,rebull_rotation_2020}.
The clearest matches came from K2 \citep{howell_k2_2014}, but a few
analogs were potentially also found using 2MASS (YLW 1C and YLW 10C;
\citealt{parks_periodic_2014}) and CoRoT
(\citealt{stauffer_csi_2015}).  We downloaded a subset of the K2 light
curves from MAST, opting for the EVEREST reductions
\citep{luger_everest_2016,luger_update_2018}.  They are plotted in
Figure~\ref{fig:brethren}.

\explain{NOTE: added citation to Bodman et al 2017 in this paragraph.}
These light curves have been phenomenologically classified as
``persistent flux dips'' or ``transient flux dips'', based on whether
their depths and durations show variability over the 90-day K2
campaigns \citep{stauffer_orbiting_2017}.  In the terminology of
\citet{stauffer_orbiting_2017}, these objects are morphologically
distinct from ``scallop shell'' light curves, and are present in stars
at more advanced evolutionary disk stages than the ``dipper'' stars
\citep{ansdell_young_2016,bodman_dippers_2017,cody_manyfaceted_2018}.
The persistent and transient flux dip stars all show angular dips that
are cannot be explained as the effects of starspots.  These stars
typically have the following things in common:
\begin{enumerate}[topsep=0.5ex,itemsep=-0.5ex,partopsep=1ex,parsep=1.5ex]
  \item They are weak-lined T Tauri stars\added{ (under the
    definition of {\it e.g.}, \citealt{fang_star_2009})}.
  \item The spectral type is M2 to M5 ({\it e.g.},
    \citealt{rebull_usco_2018},~Figure~20).
  \item The age is typically $\lesssim$ 100$\,$Myr.\footnote{ At
    present, the oldest observed ``scallops'' are in the Pleaides
    \citep{rebull_rotation_2016}. One of these, EPIC 211013604, might
    meet the ``persistent dip'' classification.  If so, it is \deleted{also} the
    oldest known. } 
  \item The light curves show shallow, angular dips, \deleted{usually}
    superposed on large-amplitude smooth variability \added{at or near
    1:1 synchronicity}. The \replaced{latter}{smooth variability} is
    interpreted as stellar rotation.
  \item The rotation is rapid, with a period that is usually between
    0.5 and 2.0 days.
  \item There is rarely any infrared excess that is detectable in the
    WISE data (never any W4 detection; only a few W3 detections).
  \item They sometimes show multiple dips per cycle.
  \item The dip depths, durations, and phases can vary over just a few
    cycles ({\it e.g.}, EPIC 204143627).
  \item The dip depths can change after flares.
  \item They are rare at a population level, with an occurrence rate
  \replaced{of $\lesssim 1\%$}{between 0.1--1\%}
   of young M2-M5 stars \citep{rebull_usco_2018}.
\end{enumerate}
The 10.76$\,$hr signal in \ptfo\ meets all of these criteria.
\replaced{This is the first connection of \ptfo\ with this class of
objects, likely because the TESS data enabled us to put the dips in
the context of the asymmetric out-of-dip modulation.} {While this was
in fact mentioned by \citet{stauffer_orbiting_2017}, the
asymmetric out-of-dip modulation shown by the TESS data solidifies the
connection.}

@@

\explain{NOTE: Removed ``high-latitude accretion hotspots'' from the
list in this paragraph}
The key question is what causes the transient dips. This is an
unsolved problem not only for \ptfo\ but also for the emerging class
of similar young and rapidly rotating M-dwarfs.  Many possible
explanations were discussed by \citet{rebull_rotation_2016},
\citet{david_transient_2017}, \citet{stauffer_orbiting_2017}, and
\citet{zhan_complex_2019}.  Among the {\it disfavored} explanations
are that the dips are caused by
{\it (i)} eclipsing binaries;
{\it (ii)} ``dipper''-flavor Class-I or Class-II disks;
{\it (iii)} eclipses of prominences;
{\it (iv)} high-latitude starspots;
or {\it (v)} dust clouds of \replaced{plausible
composition}{particular compositions
\citep[][Figure~10]{zhan_complex_2019}}.  We also view the possibility
of {\it (vi)} tidally disrupted planetary or cometary material to be
implausible, given the synchronicity between dip and rotation periods
seen across many systems.  \added{Finally, {\it (vii)} transits of an
enshrouded protoplanet can be added to the list of disfavored
explanations, based on the chromaticity, asymmetric phase curve, and
light-curve changes seen in \ptfo.}

\explain{NOTE: ``High-latitude accretion hotspots'' was moved to this
paragraph.}
The explanations that we view as not yet ruled out include {\it A)}
transiting clumps of \added{dust or }gas at the Keplerian co-rotation
radius; {\it B)} occultations of \replaced{starspots}{hotspots on the
stellar surface} by an optically thick disk \citep{zhan_complex_2019};
and {\it C)} high-latitude accretion hotspots passing behind the star
\citep{yu_tests_2015}.  \replaced{The evolution of \ptfo\
over}{Observations of \ptfo\ from} the past decade may offer some
hints.  \deleted{Specifically, }\ptfo's transition between having
none, one, and multiple dips per cycle seems important
\added{(Figure~\ref{fig:o_minus_c})}.  \deleted{The dips would then
need to be caused by material that was somehow disrupted from the
planet, but somehow remained co-orbital for an extended duration. This
seems implausible.} \added{ The H$\alpha$ excess observed by
\citet{johnskrull_h_2016} also seems like an important piece of
evidence.  }

\explain{NOTE: This paragraph is newly added, to support our view of the
high-latitude accretion hotspot being plausible for \ptfo.}
The high-latitude accretion hotspot scenario seems qualitatively
capable of explaining most of the characteristics of \ptfo, including
the fact that its H$\alpha$ equivalent width of 11.4$\,$\AA\
\citep{van_eyken_ptf_2012} places it near the ``classical/weak'' T
Tauri division \citep{briceno_cidaII_2019}, and the fact that its
excess H$\alpha$ component varies in velocity between $-200$ and
$+200$ km$\,$s$^{-1}$ \citep{johnskrull_h_2016}.  The lack of a
NIR-excess could be problematic, but does not eliminate the
possibility that a tenuous amount of dust or gas remains near the
star, and is able to accrete onto it.  We view the main problem with
this explanation to be the likelihood that it explains the population
of transient flux dips.  RIK-210 in particular has dips too deep
($\sim$20\%) and too asymmetric to be explained by occultations of a
hot spot \citep{david_transient_2017}.

\explain{NOTE: This paragraph has been expanded to clarify the
possibility of a hotspot being occulted by a disk.}
Explanations {\it A} and {\it B} (circumstellar material at the
co-rotation radius, and occultations of spots by a disk) have the
appeal that they are flexible enough to explain not only the transient
and persistent-dip stars, but also the ``scallop shells''
\citep{stauffer_orbiting_2017}.  \replaced{Despite this appeal, the
possibility of distinct mechanisms explaining these distinct
variability classes remains open.}{ In the spot occultation scenario
\citep{zhan_complex_2019}, the ``blobs'' that provide the necessary
brightness contrast are on the surface of the star, rather than at the
co-rotation radius. The observing geometry would require an inner ring
to be projected into a narrow band in front of the stellar disk, in
order to produce variations with sufficiently short timescales.  The
outer disk would also need to either not be present, or to be
misaligned enough and insufficiently flared, that it did not fall
along the line of sight.  This geometric requirement seems too
contrived to occur in even $\lesssim$1\% of young M dwarf systems. }

@@

\explain{NOTE: Updated this paragraph to remove spot occultations by a disk as
a ``preferred explanation''.}
The physical mechanism that explains the transient and persistent flux
dips is unresolved. Our preferred explanation is that a clumpy torus
of dust or gas is near the Keplerian co-rotation radius, and
periodically occults the star.  The \added{chromaticity,
out-of-eclipse variability, and} jumping orbital phase
\replaced{disfavored}{of \ptfo\ all disfavor} the explanation of an
enshrouded, transiting protoplanet.  Though \ptfo\ may not be a
planet,\deleted{ as we and others had hoped,} understanding it and its
analogs is a worthy problem\replaced{.}{ that seems likely to provide
insight into how gas and dust disperse from young protostellar disks.}
\replaced{It might even}{Understanding this process would, in turn,}
teach us about the birth environments of the majority of
habitable-zone Earth-sized planets \citep{dressing_occurrence_2013}.
